Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-26 demystify.tex
Updates from Overleaf

The most concrete evidence of research ability is having \textbf{published papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.

@@

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you should explicit say something about them in your statement.

@@

\section{Letter of Recommendation (LOR)}\label{sec:lor}

@@

However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

@@

\paragraph{Reminding Writers} After you submit your application, you can tell your writers that and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

@@

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, talk about things that adcom members might not know about, e.g., your Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants that we are considering (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

\begin{commentbox}[Hung:]
  I think asking for a “research vision” from a Ph.D. applicant is too much. Even people graduated with a Ph.D. have a great difficulty in explicating their research vision. In my opinion, you should focus on showing (not telling) your research passion (if you do have).
\end{commentbox}

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they will likely be those who interview you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part at the end of the statement, so do not skip it.

\begin{commentbox}[Vu:]
  I often read the research statement (and the LORs) first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). If I am not
  convinced, then I will likely recommend rejection (unless I see
  something standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

So what can you do about this? several things including asking your CS dept to put itself on CSRankings (it's the easiest way to get CS people to know about the school and its CS faculty)  and explaining about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS PhD programs in the US that have Vietnamese professors}.

@@

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).
So what to do with this? as mentioned in \autoref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

@@

\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which has many options for you to select from.

>>>>

The most concrete evidence of research ability is having \textbf{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Students  accepted to top schools often have multiple first-authored papers at top venues.

@@

However, it is understandable that many students do not have the opportunities to publish papers. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

@@

\section{Letters of Recommendation (LOR)}\label{sec:lor}

@@

However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS). Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potentials. Letters focusing your course work or non-research projects at a company won't carry much weight.

@@

\paragraph{Reminding Writers} After you submit your application, you should tell your writers about that and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

@@

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. Also, talk about things that adcom members might not know about, e.g., your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

% \begin{commentbox}[Hung:]
%   I think asking for a “research vision” from a Ph.D. applicant is too much. Even people graduated with a Ph.D. have a great difficulty in explicating their research vision. In my opinion, you should focus on showing (not telling) your research passion (if you do have).
% \end{commentbox}

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part at the end of the statement, so do not skip it.

\begin{commentbox}[Vu:]
  I often read the research statement and LORs first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

So what can you do about this? You can explain about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS PhD programs in the US that have Vietnamese professors}. It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).
So what to do with this? As mentioned in \autoref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

@@

\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which is required for assistantships, e.g., TA~\autoref{sec:funding}.

### 2024-02-26 faculty.tex
Updates from Overleaf

\mainmatter

@@

>>>>

\mainmatter
\chapter{Intro}

\section{The 5 jobs of an assistant professor}
At an R1 university, an assistant professor in CS (and likely in other fields) typically has \emph{five} jobs:

\begin{enumerate}
\item[Research] You need to do research and publish papers.  To establish yourself as an independent research, you should publish papers with your students (and not solely with your PhD or postdoc adviser).

\item[Funding] You need to write proposal and secure funding from sponsors including government agencies (e.g., NSF) or industry.

\item[Mentoring] You need to recruit, mentor, and graduate PhD students.

\item[Teaching] You need to teach!

\item[Service]  You need to serve for your professional research communities (e.g., PC, review journals, NSF panels) and for your institution (e.g., PhD admission committees, dissertation committee).  You also need to do various other things including writing LoRs for your students (and even your peers)

\end{enumerate}

So, there you go. A new assistant prof. opening their own lab starting their academic career typically has these 5 duties.  In some sense having your own lab is like opening your own start up company. You have to work on raising funding, recruiting employees, do the actual work, etc.  You don't have a boss, but you are responsible for everything.

\section{Pros and Cons}

Pros:
\begin{itemize}
\item Freedom. You don’t have a boss. Seriously, you don't.  You can work on any research topics you want (preferably something that can get grants), choose your own students (not that easy to find good students) and collaborators, work on your own time, and have plenty of free time(technically I only have to be "at work" for 3 hrs a week for my class from 4:15PM--7:10PM).

\item Students. It's a great award when seeing how your students evolve to be more mature and become an expert in their research.

\item Flexibility. You can do pretty much anything.  Travels, give talks, write paper/blogs/books (like this one).  Create new courses and even online courses.  Collaborate  with academia, government, research lab, industry.  Start your own company when you retire!

\item Time with family. This might seem like a weird pro, but as a professor, I have much more
control over my time than many other jobs. It lets me schedule my work so that I can
spend more with my family. For example, when my ﬁrst kid was born, my department
gave me teaching time off so that I could spend time with him. I spent six months with
my son – irreplaceable time that is harder to get in many other jobs (this is especially
true in the USA where paternity and maternity leave is minimal in many jobs).
\end{itemize}

Cons:
\begin{itemize}
\item Responsibility: The role carries high responsibility, not only for one's own career but also for the careers of PhD students under their supervision.
\item Stress: The initial years are particularly stressful as one learns various aspects of the job while under pressure.
\item Poor financial compensation: Assistant professors typically earn less than they would in industry, leading to a significant opportunity cost over time.
\item Funding: Securing research funds requires substantial time and effort, with no guarantee of success, given the competitive nature of grant applications.
\item Culture of unpaid labor: Academia in the USA often relies on unpaid work, both from faculty and students, which can be a significant drawback compared to industry where compensation is typically fairer.
\end{itemize}

@@

https://vijay03.github.io/asstprofbook/chapters/job.pdf
